---
title: "Project 2"
author: "Ritesh Bachar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Fun with Functions

```{r part1A Exponential transformation}
# creating the function to compute exponential using series expansion
exp <- function(x, k) {
 
  # Adding the terms of the series expansion up to k terms
  for (i in 1:k) {
    result <- 1 + x + (x^i) / factorial(i)
  }
  
  return(result)
}

# Example of using function
exp(4, 6) 

```

```{r part1B Sample mean and sample standard deviation}
#writing the function for calculating mean
sample_mean <- function(x) {
  {N <- length(x)
   avg <- sum(x)/N
  }
  return(avg)
}

example_1B <- c(1, 3, 5, 7, 9)
#example of using mean function
sample_mean(example_1B)

#writing the function for calculating standard deviation
sample_sd <- function(x) {
  {
    N <- length(x)
    avg <- sample_mean(x)
    sum_of_squares <- sum((x-avg)^2)
    sd <- sqrt(1/(N-1)*sum_of_squares)
  }
  return(sd)
}

#example of using std deviation function
sample_sd(example_1B)

```

```{r part1C Confidence Intervals}
# Creating function to calculate confidence interval
calculate_CI <- function(x, conf = 0.95) {
  N <- length(x)  # the number of observations
  mean_x <- sample_mean(x)  # using sample_mean functions from before
  sd_x <- sample_sd(x)  # using sample_sd functions from before
  SE <- sd_x / sqrt(N)  # Standard error of the mean
  
  alpha <- 1 - conf  # alpha value
  degrees_freedom <- N - 1  # degrees of freedom for the t-distribution
  t_score <- qt(p = alpha / 2, df = degrees_freedom, lower.tail = FALSE)  # t-score
  
  # confidence interval
  margin_of_error <- t_score * SE
  lower_bound <- mean_x - margin_of_error
  upper_bound <- mean_x + margin_of_error
  
  # Return the confidence interval as a named vector
  return(c(lower_bound = lower_bound, upper_bound = upper_bound))
}

# Example usage:
example_1C <- c(11, 13, 15, 17, 19)

# Calculate 95% confidence interval
CI_95 <- calculate_CI(example_1C, conf = 0.95)
print (CI_95)

#Checking if my CI function matches with existing function in R

dat <- data.frame(x = example_1C)
fit <- lm(x ~ 1, dat)

# Calculate a 95% confidence interval
confint(fit, level = 0.95)

```

## Part 2: Wrangling Data

``` {r part2 wrangling data} 
#loading tidytuesday
library(here)
#creating data folder
if (!dir.exists(here("data"))) {
    dir.create(here("data"))
}

#loading tidytuesday data
if (!file.exists(here("data", "tuesdata_rainfall.RDS"))) {
    tuesdata <- tidytuesdayR::tt_load("2020-01-07")
    rainfall <- tuesdata$rainfall
    temperature <- tuesdata$temperature

    # save the files to RDS objects
    saveRDS(tuesdata$rainfall, file = here("data", "tuesdata_rainfall.RDS"))
    saveRDS(tuesdata$temperature, file = here("data", "tuesdata_temperature.RDS"))
}

#loading the datasets
rainfall <- readRDS(here("data", "tuesdata_rainfall.RDS"))
temperature <- readRDS(here("data", "tuesdata_temperature.RDS"))

library(tidyverse)

#taking a glimpse of the 2 datasets
glimpse(rainfall)
glimpse(temperature)

#Load Libraries
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)

#dropping rows with NAs in rainfall
rainfall_without_na <- rainfall %>% drop_na()

#creating new date column (year+month+day)
#creating new column with year only

rainfall_without_na <- rainfall_without_na %>%
  mutate(
    date = ymd(paste(year, month, day, sep = "-")),
      year_only = year 
)

#dropping the year, month, and day columns
rainfall_without_na <- rainfall_without_na %>%
  select(-year, -month, -day)

class(rainfall_without_na$date)
class(rainfall_without_na$year_only)

#converting city_name to uppercase
rainfall_without_na <- rainfall_without_na %>%
  mutate(city_name = str_to_upper(city_name))

#joining rainfall and temp datasets using innerjoin
#to keep only obs present in both datasets

df <- rainfall_without_na %>%
  inner_join(temperature, by = c("date" = "date", "city_name" = "city_name"))

#printing results of join
as_tibble(df)

```

## Part 3: Data Visualization

```{r Part 3A: Plotting temperature data over time}
library(ggplot2)
#considering data for years 2014 and onwards
df_after_2014 <- df %>%
  filter(year_only >= "2014")

#city min max temp in the same plot with 2 different colors
temp_over_time <- ggplot( data = df_after_2014, mapping = aes(x = date, y = temperature)) +
    geom_line( aes(color = temp_type) ) +
    facet_wrap(~ city_name, scales = "free_y") + 
    ylim(-10, 50) +
    scale_color_manual(values = c("min" = "green4", "max" = "brown4"),
                       labels = c( "min" = "Minimum Temperature", "max" = "Maximum Temperature") ) + 
    labs(
    title = "Temperatures Trends in Australian Cities (2014 and onwards)",
    subtitle = "Temperature variability in Canberra is the highest compared to the other cities",
    x = "Years",
    y = "Temperature (°C)",
    color = "Temperature type",
    caption = "Data source: TidyTuesday") +
    theme_gray() +
    theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.text = element_text(face = "bold"),
    legend.position = c(1, 0), 
    legend.justification = c(1, 0) 
  )

#printing the plot
print(temp_over_time)

```

```{r Part 3B: Plotting rainfall over time}
plot_rainfall <- function(city_input, year_input) {
  
  # Filter the data for the given city and year
  df_city_year <- df %>%
    filter(city_name == city_input, year_only == year_input)
  
  # Error handling if no data exists for the input city and year
  if (nrow(df_city_year) == 0) {
    stop("No data available for the specified city and year.")
  }
  
  # Log-transform the rainfall data (log1p to handle zero rainfall values)
  df_city_year <- df_city_year %>%
    mutate(log_rainfall = log1p(rainfall))
  
  # Create a histogram of log-transformed rainfall
  ggplot(df_city_year, aes(x = log_rainfall)) +
    geom_histogram(bins = 25, fill = "blue", color = "black") +
    labs(
      title = paste("Rainfall Trend for", city_input, "in", year_input),
      subtitle = paste("Frequency of Rainfall in log scale for", city_input, "in", year_input),
      x = "Rainfall in log scale",
      y = "Frequency",
      caption = "Data source: TidyTuesday"
    ) +
     theme_gray() +
    theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.text = element_text(face = "bold")
  )
}

# Example of the plot function
plot_rainfall("PERTH", 1980)

```

### Description of designing the function
I chose to input city and year and be able to see the log transformed rainfall for that city in that year. For this, I first filtered the **df** data frame. As all city and year did not have all measurements so, I created a *stop* function that would mention about missing values rather than showing an error. This would help to understand about the missing values. After that, I included **ggplot** in the function and used the city and year inputs and the log transformed rainfall to give an output of histogram. When this function is called with different city names and year, the histogram will show the log rainfall data of that specific city in that specific year. Finally, I customized the title, subtitle, axis labels for better viewing and understanding of the plot.


## Part 4: Apply functions and plot

```{r Part 4A: Tasks}
#filtering years 2014 and onwards
df_after_2014 <- df %>%
  filter(year_only >= "2014")

#grouping by city and year, and calculating the statistics
rain_df <- df_after_2014 %>%
  group_by(city_name, year_only) %>%
  summarise(
    mean_rainfall = sample_mean(rainfall),
    sd_rainfall = sample_sd(rainfall),
    
    # Using calculate_CI and assigning both lower and upper bounds at once
    CI = list(calculate_CI(rainfall)),  # Storing the CI as a list temporarily
    lower_bound = CI[[1]]["lower_bound"],
    upper_bound = CI[[1]]["upper_bound"]
  ) %>%
  ungroup() %>%
  select(-CI)  # Removing the temporary CI list column

# View the summarized data
print(rain_df)

```

```{r part 4B plotting the rain_df }
# Creating the final plot
final_plot <- ggplot(data = rain_df, aes(x = year_only)) +
  geom_point(aes(y = mean_rainfall), color = "blue") +  # geom points for mean rainfall
  geom_line(aes(y = mean_rainfall), color = "blue") +   # geom line connecting mean points
  geom_errorbar(aes(ymin = lower_bound, ymax = upper_bound), width = 0.2) +  # error bars for CI
  facet_wrap(~ city_name, scales = "free_y") +  # faceted by city with free y-axis scales
    ylim(0, 11) + #inserting ylim to better see the yaxis
  labs(
    title = "Rainfall trends in Australian Cities (2014 and beyond)",
    subtitle = "On average Perth has the most showers in a year followed by Melbourne and Brisbane,\nCanberra and Sydney showers are similar over the years",
    x = "Year",
    y = "Average Rainfall (mm)",
    caption = "Data Source: TidyTuesday"
  ) +
  theme_gray() +
    theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 10, face = "bold"),
    strip.text = element_text(face = "bold")
  )

#printing the final plot
print(final_plot)

```

